---
title: "The idlBNs Package"
author:
- name: Robert Castelo
  affiliation:
  - &id Dept. of Medicine and Life Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: robert.castelo@upf.edu
package: "`r packageDescription('idlBNs')$Version`"
abstract: >
  The `idlBNs` package implements inclusion-driven algorithms for learning the structure of Bayesian networks. It currently provides the hill-climber Monte Carlo (HCMC) algorithm for observational Gaussian data, and the interventional HCMC (iHCMC) for interventional Gaussian data.
vignette: >
  %\VignetteIndexEntry{The idlBNs Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: true
    fig_captions: yes
bibliography: bibliography.bib
---

```{r setup, echo=FALSE}
library(knitr)

options(width=80)

knitr::opts_chunk$set(
    collapse=TRUE,
    comment="")
```

# Introduction

Bayesian networks are multivariate models that assume that their joint
probability distribution factorizes according to an acyclic directed graph (DAG).
The DAG structure of a Bayesian network can be estimated from data, and this is
the so-called problem of *structure learning*. The `idlBNs` package implements
algorithms for learning the structure of Bayesian networks from observational
and interventional Gaussian data.

# Inclusion-driven structure learning

Structure learning algorithm for Bayesian networks can be classified into three
main categories: constraint-based, score-based, and hybrid algorithms. The
`idlBNs` package implements score-based algorithms that attempt following an
inclusion path in the search space of DAGs [@KocBou2k1]. Following an inclusion
path confers makes the algorithm consistent under the faithfulness assumption,
that is, it converges to the generative structure as the sample size grows large
[@Chi02; @CasKoc2k3].

Inclusion-driven learning of Bayesian networks, or *idlBNs* for
short^[Pronounced *ideal BNs*], can be implemented either in the space of the
canonical elements of the Markov equivalence classes of DAGs, also known as
*essential graphs*, as in the Greedy Equivalence Search (GES) algorithm
[@Chi02], or in the space of DAGs, as in the Hill-Climber Monte Carlo (HCMC)
algorithm [@KocCas2k1; @CasKoc2k3]. The GES algorithm was adapted to
interventional data in the Greedy Interventional Equivalence Search (GIES)
algorithm [@HauBuh12], and an efficient implementation of both is available in
the [pcalg](https://cran.r-project.org/package=pcalg) package [@KalBuh07],
through the functions `ges()` and `gies()`, respectively.

The `idlBNs` provides an implementation in R of the HCMC algorithm for
observational Gaussian data, and of the interventional HCMC (iHCMC) algorithm
for a mixture of observational and interventional Gaussian data [@Cas26]
through the function `hcmc()`.

# Scoring functions for observational and interventional Gaussian data

The `idlBNs` package implements two different functions for scoring a DAG
structure from observational and interventional Gaussian data. The `iBIC()`
function, which implements the interventional Bayesian Information Criterion
(iBIC) score [@HauBuh12; @HauBuh15], and the `iBGe()` function, which implements
the interventional Bayesian Gaussian equivalent (iBGe) score [@KuiMof25].

# How to use the idlBNs package

Here we illustrate how to use the idlBNs package with multivariate Gaussian
data, simulated using the procedures of the
[pcalg](https://cran.r-project.org/package=pcalg) package for that purpose.
Besides this package, the idlBNs package currently depends also on the
Bioconductor package [graph](https://bioconductor.org/packages/graph) for
building and representing DAGs through the object class `graphNEL`.

```{r, message=FALSE}
library(graph)
library(pcalg)
library(idlBNs)
```
We start by simulating a Gaussian Bayesian network with 10 vertices and 50%
density using the function `r.gauss.pardag()` from the `pcalg` package,
extracting its DAG strcture and converting it into an essential graph using
the function `dag2essgraph()` from the `pcalg`  package.

```{r}
p <- 5
d <- 0.6
set.seed(123)
M_G <- r.gauss.pardag(p, d, top.sort=TRUE, normalize=TRUE)
G <- as(M_G, "graphNEL")
E <- dag2essgraph(as(G, "graphNEL"))
```
Assume we want to consider two different singleton intervention targets ($k=2$),
i.e., we select two vertices from the DAG structure uniformly at random to
intervene in their associated random variables, but we intervene only in one
random variable at a time in a given multivariate observation. Following the
terminology of @HauBuh12, we define a family of *intervention targets*
$\mathcal{I} = \{I_1, I_2, I_3\}$, where $I_1=\{\emptyset\}$ and $I_i\{v_i\}$
with $i=\{2, 3\}$ and $v_i\in V$ is a randomly selected vertex from the DAG
structure $G=(V, E)$.

```{r}
k <- 2
I <- c(list(integer(0)), sample(p, size=k, replace=FALSE))
I
```
Consider simulating a sample size of $n=100$ multivariate observations, with a
similar number of them for each intervention target.

```{r}
n <- 30
nbytgts <- rep(floor(n / (k + 1)), k)
nbytgts <- c(n - sum(nbytgts), nbytgts)
nbytgts
```
Simulate mixed observational and interventional multivariate Gaussian data from
the DAG structure $G$ and the family of intervention targets $\mathcal{I}$ using
the function `rmvnorm.ivent()` from the `pcalg` package, storing the index to
the corresponding intervention target for each row of the data.

```{r}
dat <- list()
for (v in seq_along(I)) {
  targets <- I[[v]]
  dat[[v]] <- rmvnorm.ivent(nbytgts[v], M_G, target=targets,
                            target.value=rep(2, length(targets)))
}
dat <- do.call("rbind", dat)
tindex <- rep(1:length(nbytgts), nbytgts)
```
Run the iHCMC algorithm informing the presence of interventional data. By default,
the iBIC score is used in the search algorithm.

```{r, message=FALSE}
Dhat.ihcmc <- hcmc(dat, targets=I, target.index=tindex)
Dhat.ihcmc
```
Calculate the structural Hamming distance (SHD) to the essential graph of the
generative DAG structure $G$ using the function `shd()` from the `pcalg`
package.

```{r}
shd(dag2essgraph(Dhat.ihcmc$dag), E)
```
Repeat the search using the iBGe score instead of the iBIC score.

```{r, message=FALSE}
Dhat.ihcmc.ibge <- hcmc(dat, targets=I, target.index=tindex, scorefun=iBGe)
shd(dag2essgraph(Dhat.ihcmc.ibge$dag), E)
```
We can see that the iBGe score leads to an estimate of the DAG structure that
is closer to the generative DAG structure, in terms of the SHD. Now use the
HCMC algorithm without informing the presence of interventional data, i.e.,
assuming that all data are observational.

```{r, message=FALSE}
Dhat.hcmc <- hcmc(dat)
shd(dag2essgraph(Dhat.hcmc$dag), E)
Dhat.hcmc.ibge <- hcmc(dat, scorefun=iBGe)
shd(dag2essgraph(Dhat.hcmc.ibge$dag), E)
```
In this setting, using either the BIC or the BGe score leads to worse estimates of
the DAG structure.

# Computational performance benchmark

Currently, the iBIC and iBGe scores, as well as the iHCMC algorithm, are
implemented using only the R language without any kind of optimization. In
practice, this means the iHCMC algorithm can only be applied to a handful of
vertices and this implementation is currently only useful for research
prototyping and educational purposes. Using the CRAN package
[bench](https://cran.r-project.org/package=bench) [@HesVau25], here we
benchmark the computational performance of the iHCMC algorithm for two
different numbers of vertices, a given sample size and two randomly selected
intervention targets, to keep track of how its performance improves as we
optimize its implementation through the next versions of the `idlBNs` package.

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
library(dplyr)
library(tibble)
library(bench)
library(cli)

if (Sys.getenv("IDLBNS_BENCHMARK") == "TRUE") {
    if (Sys.getenv("IDLBNS_BENCHMARK_RESULTS_PATH") != "")
        if (!dir.exists(Sys.getenv("IDLBNS_BENCHMARK_RESULTS_PATH"))) {
            msg <- "The path in IDLBNS_BENCHMARK_RESULTS_PATH does not exist."
            cli_abort(c("x"=msg))
        }

    bmdat <- list()
    p <- c(10, 20)
    n <- 100
    for (i in seq_along(p)) {
        D <- r.gauss.pardag(p[i], 0.2, top.sort=TRUE, normalize=TRUE)
        I <- list(integer(0), sample(p[i], size=2, replace=FALSE))
        nbytgts <- c(n/2, n/2)    
        dat <- rbind(rmvnorm.ivent(nbytgts[1], D, target=I[[1]],
                                   target.value=rep(2, length(I[[1]]))),
                     rmvnorm.ivent(nbytgts[2], D, target=I[[2]],
                                   target.value=rep(2, length(I[[2]]))))
        tindex <- rep(1:length(nbytgts), nbytgts)
        bmdat[[i]] <- bench::mark(iBIC=hcmc(dat, targets=I, target.index=tindex,
                                            scorefun=iBIC, verbose=FALSE),
                                  iBGe=hcmc(dat, targets=I, target.index=tindex,
                                            scorefun=iBGe, verbose=FALSE),
                                  iterations=3, check=FALSE)
    }

    bmdat <- do.call("rbind", lapply(bmdat,
                                     function(x) x[, c("expression", "median",
                                                       "mem_alloc")]))
    bmdat <- rename(bmdat, score_function=expression)
    bmdat$score_function <- attr(bmdat$score_function, "description")
    bmdat <- bmdat |> add_column(p=rep(p, each=length(p)), .before="score_function")
    bmdat <- bmdat |> add_column(version=packageVersion("idlBNs"), .before="p")
    if (Sys.getenv("IDLBNS_BENCHMARK_RESULTS_PATH") != "") {
        fname <- file.path(Sys.getenv("IDLBNS_BENCHMARK_RESULTS_PATH"),
                           "benchmark_results.csv")
        bmdat2 <- bmdat
        bmdat2$version <- as.character(bmdat2$version)
        bmdat2$median <- as.numeric(bmdat2$median)
        bmdat2$mem_alloc <- as.numeric(bmdat2$mem_alloc)
        write.csv(bmdat2, file=fname, row.names=FALSE)
    }
}
```
```{r, eval=FALSE}
library(dplyr)
library(tibble)
library(bench)

bmdat <- list()
p <- c(10, 20)
n <- 100
for (i in seq_along(p)) {
    D <- r.gauss.pardag(p[i], 0.2, top.sort=TRUE, normalize=TRUE)
    I <- list(integer(0), sample(p[i], size=2, replace=FALSE))
    nbytgts <- c(n/2, n/2)    
    dat <- rbind(rmvnorm.ivent(nbytgts[1], D, target=I[[1]],
                               target.value=rep(2, length(I[[1]]))),
                 rmvnorm.ivent(nbytgts[2], D, target=I[[2]],
                               target.value=rep(2, length(I[[2]]))))
    tindex <- rep(1:length(nbytgts), nbytgts)
    bmdat[[i]] <- bench::mark(iBIC=hcmc(dat, targets=I, target.index=tindex,
                                        scorefun=iBIC, verbose=FALSE),
                              iBGe=hcmc(dat, targets=I, target.index=tindex,
                                        scorefun=iBGe, verbose=FALSE),
                              iterations=3, check=FALSE)
}

bmdat <- do.call("rbind", lapply(bmdat, function(x) x[, c("expression", "median", "mem_alloc")]))
bmdat <- rename(bmdat, score_function=expression)
bmdat$score_function <- attr(bmdat$score_function, "description")
bmdat <- bmdat |> add_column(p=rep(p, each=length(p)), .before="score_function")
bmdat <- bmdat |> add_column(version=packageVersion("idlBNs"), .before="p")
```
Table \@ref(tab:benchmark-results) below shows the benchmarking results of this
version of the `idlBNs` package, jointly with the results of previous versions,
if available, for comparison purposes. Actually, if you want to run the benchmark
when building the vignette, you should set the environment variable
`IDLBNS_BENCHMARK=TRUE` before building it, and if you want to save the results
of the benchmark, you should set the environment variable
`IDLBNS_BENCHMARK_RESULTS_PATH` to a valid path where the results will be saved
in a CSV file named `benchmark_results.csv`.

```{r benchmark-results, message=FALSE, echo=FALSE, results='asis'}
library(kableExtra)

fname <- system.file("extdata", "benchmark_results.csv", package="idlBNs")
if (file.exists(fname)) {
    prevbmdat <- read.csv(fname)
    prevbmdat$version <- package_version(prevbmdat$version)
    prevbmdat$median <- as_bench_time(prevbmdat$median)
    prevbmdat$mem_alloc <- as_bench_bytes(prevbmdat$mem_alloc)
    if (Sys.getenv("IDLBNS_BENCHMARK") == "TRUE") {
      if (!bmdat$version[1] %in% prevbmdat$version)
          bmdat <- rbind(prevbmdat, bmdat)
    } else
      bmdat <- prevbmdat
}

kable(bmdat, format="html",
      col.names=c("Version", "Vertices", "Score Function", "Median Time",
                  "Memory Consumption"),
      caption="Benchmark of computational performance of the iHCMC algorithm implemented in `hcmc()` for p={10, 20} vertices, and the score functions `iBIC()` and `iBGe()`.") |>
      kable_styling(bootstrap_options=c("striped", "hover", "condensed"),
                  full_width=FALSE, position="center")
```




# Session information

```{r session_info, cache=FALSE}
sessionInfo()
```

# References
